Can the symmetric \textit{Fermi} and eROSITA bubbles be produced by tilted jets: a proof-of-concept study

### 2022-03-24 paper.tex
update

% Abstract of the paper
\begin{abstract}
The \textit{Fermi Gamma-Ray Space Telescope} reveals two large bubbles in the Galaxy,\
which extend nearly symmetrically $\sim50^{\circ}$ above and below the Galactic center (GC).\
The recent discovery of giant eROSITA bubbles also shows such a symmetry about the GC, suggesting that they may originate from a single GC activity.\
Previous simulations of bubble formation that invoke active galactic nucleus (AGN) jets\
have assumed that the jets are vertical to the Galactic plane;\
however, in general there does not need to be a correlation between the jet orientation and the rotational axis of the Galactic disk.\
Using three-dimensional (3D) special relativistic hydrodynamic (SRHD) simulations that\
include cosmic rays (CRs) and thermal gas, we show that the interaction between the AGN jets and the thin interstellar medium (ISM) disk is crucial for producing the symmetric \textit{Fermi} and eROSITA bubbles (collectively, Galactic bubbles).\
%We also produce simulated gamma-ray and\
%microwave spectra from the inverse Compton (IC) scattering and synchrotron,\
%and compare them with the observed counterparts in order to verify\
%an oblique AGN jets model.
Specifically, we find that\
(1) the simulated Galactic bubbles are nearly symmetric about the GC albeit\
the bipolar jets are at an angle $45^{\circ}$ with respect to\
the rotational axis of the Galaxy;\
(2) the edge of the eROSITA bubbles corresponds to a forward shock,\
originally driven by oblique bipolar jets emanating from the GC 12.39 million years ago; %,\
%and significantly stretched by the stratified atmosphere afterwards.\
(3) followed by the forward shock is a tangling contact discontinuity,\
which we interpret as the edge of the \textit{Fermi} bubbles\
composed of turbulent and high-temperature ($\sim2$ keV)\
plasma in pressure balance with the external medium;
(4) assuming a leptonic model with spatially uniform CR electron spectrum, we find that the observed gamma-ray and microwave spectra can be reproduced with a best-fit CR power-law index of 2.4.\
The broad agreement between simulated and observed multi-wavelength features suggests that forming the Galactic bubbles by oblique AGN jets is a plausible scenario.
%has thus released a caveat:\
%the jets shall be vertical to the disc normal.
\end{abstract}

@@

\section{Introduction}
The detection of the \textit{Fermi} bubbles \citep{Su2012,Ackermann2014,Narayanan2017},\
two large bubbles symmetrically extending about 50 degrees above and below the Galactic plane,\
is one of the great discoveries of the \textit{Fermi} Large Area Telescope \citep{Atwood2009}.\
The gamma-ray emission of the \textit{Fermi} bubbles is observed in the energy range\
of $\sim$1--100 GeV and has an almost spatially uniform hard\
spectrum, sharp edges and an approximately flat brightness distribution (see \cite{Yang2018} for a review). Recently, the newly launched eROSITA \citep{Predehl2021} conducted an all-sky X-ray survey with high-spatial resolution and revealed two gigantic bubbles (eROSITA bubbles hereafter) extending to $\sim 80$ degrees in Galactic latitudes, corresponding to an intrinsic size of 14 kpc across \citep{Predehl2020}.\
The remarkable resemblance between the eROSITA and \textit{Fermi} bubbles suggest that they likely share the same origin \citep{Yang2022}.
Their symmetry about the GC further suggests that these Galactic bubbles may be generated by powerful energy injections from the GC, possibly related to nuclear star formation\
\citep{PhysRevLett.106.101102,Carretti2013,Crocker2015,Sarkar2015}
or past AGN activity \citep{Guo2012,Guo2012b,Yang2012,Yang2013,Mou2014,Yang2017}. The latter scenario is what we will focus on in this work.

@@

  We use the following quantities to characterize the jets:
  the density contrast between the thermal gas contained in the jet source and the ambient gas,\
  $\rho_{\text{jet}}/\rho_{\text{amb}}=10^{-3}$,\
  the temperature contrast, $T_{\text{jet}}/T_{\text{amb}}=2\times10^{4}$,\
  the CR-to-gas pressure ratio of 0.18,\
  and the flow 4-velocity inside the jet source along the jet axis, $\beta\gamma = 0.6$.\
  The jet power is thus $3.2\times 10^{42}$ erg s$^{-1}$, resulting in an Eddington ratio of 0.008.\
  Note that since we inject the jets at the center of the clumpy disk,\
  we define the atmosphere gas density by the peak density of\
  the isothermal disc on the mid-plane $z=0$ (i.e. $\rho^{\text{peak}}_{\text{isoDisk}}$),\
  as opposed to the \textit{clumpy} density around the jet source.

@@

 \begin{figure}%
      \captionsetup[subfigure]{labelformat=simple}
      \begin{subfigure}[b]{0.2\linewidth}%
         \includegraphics[height=10cm]{figures/fig__jetI5+ismSeed3-45deg-a}%
         \caption{}%
         \label{fig__jetI5+ismSeed3-45deg-a}%
      \end{subfigure}%
      \hspace{10pt}%
      \begin{subfigure}[b]{0.2\linewidth}%
         \includegraphics[height=10cm]{figures/fig__jetI5+ismSeed3-45deg-b}%
         \caption{}%
         \label{fig__jetI5+ismSeed3-45deg-b}%
      \end{subfigure}%
      \hspace{-2pt}%
      \begin{subfigure}[b]{0.2\linewidth}%
         \includegraphics[height=10cm]{figures/fig__jetI5+ismSeed3-45deg-c}%
         \caption{}%
         \label{fig__jetI5+ismSeed3-45deg-c}%
      \end{subfigure}%
      \hspace{-2pt}%
      \begin{subfigure}[b]{0.2\linewidth}%
         \includegraphics[height=10cm]{figures/fig__jetI5+ismSeed3-45deg-d}%
         \caption{}%
         \label{fig__jetI5+ismSeed3-45deg-d}%
     \end{subfigure}%
      \caption{
             The slices of pressure (top), temperature (middle), and number density (bottom)\
             at the end of simulation $t=12.39$ Myr.\
             The slices pass through a bipolar jet source injecting along $z=-y$ direction\
             for a duration $t=0$--$1.2$ Myr.\
             Comparison between the clumpy (Fig. \ref{fig__jetI5+ismSeed3-45deg-a})\
             and the smooth disks (Fig. \ref{fig__jetI5+ismSeed3-45deg-b}) in a stratified atmosphere\
             shows that the initial density distribution of the dense disk has an insignificant effect\
             on the overall dynamics of bubbles. However, the outermost bubbles arising from the smooth disc\
             in an uniform atmosphere (Fig. \ref{fig__jetI5+ismSeed3-45deg-c}) is nearly spherical,\
             suggesting that the stratification facilitates the elongation of the outermost bubbles significantly.\
             Also, Fig. \ref{fig__jetI5+ismSeed3-45deg-c} and \ref{fig__jetI5+ismSeed3-45deg-d}\
             reveal that the development of the innermost bubbles\
             is always associated with the disk. Also, without the disk (Fig. \ref{fig__jetI5+ismSeed3-45deg-d}),\
             the outermost bubbles and the turbulent plasma would be oblique,
             indicating that the dense disk is crucial for the production of symmetric Galactic bubbles.\ {\color{red} KY: Consider reducing the caption as it is essentially the same as the text.}
      }
      \label{fig__jetI5+ismSeed3-45deg}
 \end{figure}%

@@

\begin{figure*}
  \includegraphics[width=\linewidth]{figures/fig__GammaRay_100e9_1e6_angle_000.png}
  \caption{The observed (left; \citealt{Selig2015}) and simulated (right) photon flux\
           in the energy bin $76.8-153.6$ GeV.\
           Note that the left panel is the\
           photon flux of the diffuse component reconstructed by the D$^3$PO\
           algorithm \citep{Selig2015} that analyzes\
           the photon data from the \textit{Fermi} Large Area Telescope \citep{Atwood2009}\
           and removes the contribution from point-like component.
           The red arrow at the center of the right\
           panel depicts the direction of the bipolar jet, constantly ejecting at an angle of $45^{\circ}$\
           to the disc normal in 1.2 Myr.
  }
  \label{fig__gammaRay-map}
\end{figure*}

>>>>

% Abstract of the paper
\begin{abstract}
The \textit{Fermi Gamma-Ray Space Telescope} reveals two large bubbles in the Galaxy,\
which extend nearly symmetrically $\sim50^{\circ}$ above and below the Galactic center (GC).\
The recent discovery of giant eROSITA bubbles also shows such a symmetry about the GC, suggesting that they may originate from a single GC activity.\
Previous simulations of bubble formation that invoke active galactic nucleus (AGN) jets\
have assumed that the jets are vertical to the Galactic plane;\
however, in general there does not need to be a correlation between the jet orientation and the rotational axis of the Galactic disk.\
Using three-dimensional (3D) special relativistic hydrodynamic (SRHD) simulations that\
include cosmic rays (CRs) and thermal gas, we show that the interaction between the AGN jets and the thin interstellar medium (ISM) disk is crucial for producing the symmetric \textit{Fermi} and eROSITA bubbles (collectively, Galactic bubbles).\
%We also produce simulated gamma-ray and\
%microwave spectra from the inverse Compton (IC) scattering and synchrotron,\
%and compare them with the observed counterparts in order to verify\
%an oblique AGN jets model.
Specifically, we find that\
(1) the simulated Galactic bubbles are nearly symmetric about the GC albeit\
the bipolar jets are at an angle $45^{\circ}$ with respect to\
the rotational axis of the Galaxy;\
(2) the edge of the eROSITA bubbles corresponds to a forward shock,\
originally driven by oblique bipolar jets emanating from the GC 12.39 million years ago; %,\
%and significantly stretched by the stratified atmosphere afterwards.\
(3) followed by the forward shock is a tangling contact discontinuity,\
which we interpret as the edge of the \textit{Fermi} bubbles\
composed of turbulent and high-temperature ($\sim2$ keV)\
plasma in pressure balance with the external medium;
(4) assuming a leptonic model with spatially uniform CR electron spectrum, we find that the observed gamma-ray and microwave spectra can be reproduced with a best-fit CR power-law index of 2.4.\
The broad agreement between simulated and observed multi-wavelength features suggests that forming the Galactic bubbles by oblique AGN jets is a plausible scenario.
%has thus released a caveat:\
%the jets shall be vertical to the disk normal.
\end{abstract}

@@

\section{Introduction}
The detection of the \textit{Fermi} bubbles \citep{Su2012,Ackermann2014,Narayanan2017},\
two large bubbles symmetrically extending about 50 degrees above and below the Galactic plane,\
is one of the great discoveries of the \textit{Fermi} Large Area Telescope \citep{Atwood2009}.\
The gamma-ray emission of the \textit{Fermi} bubbles is observed in the energy range\
of $\sim$1--100 GeV and has an almost spatially uniform hard\
spectrum, sharp edges and an approximately flat brightness distribution (see \citealt{Yang2018} for a review). Recently, the newly launched eROSITA \citep{Predehl2021} conducted an all-sky X-ray survey with high-spatial resolution and revealed two gigantic bubbles (eROSITA bubbles hereafter) extending to $\sim 80$ degrees in Galactic latitudes, corresponding to an intrinsic size of 14 kpc across \citep{Predehl2020}.\
The remarkable resemblance between the eROSITA and \textit{Fermi} bubbles suggest that they likely share the same origin \citep{Yang2022}.
Their symmetry about the GC further suggests that these Galactic bubbles may be generated by powerful energy injections from the GC, possibly related to nuclear star formation\
\citep{PhysRevLett.106.101102,Carretti2013,Crocker2015,Sarkar2015}
or past AGN activity \citep{Guo2012,Guo2012b,Yang2012,Yang2013,Mou2014,Yang2017}. The latter scenario is what we will focus on in this work.

@@

  We use the following quantities to characterize the jets:
  the density contrast between the thermal gas contained in the jet source and the ambient gas,\
  $\rho_{\text{jet}}/\rho_{\text{amb}}=10^{-3}$,\
  the temperature contrast, $T_{\text{jet}}/T_{\text{amb}}=2\times10^{4}$,\
  the CR-to-gas pressure ratio of 0.18,\
  and the flow 4-velocity inside the jet source along the jet axis, $\beta\gamma = 0.6$.\
  The jet power is thus $3.2\times 10^{42}$ erg s$^{-1}$, resulting in an Eddington ratio of 0.008.\
  Note that since we inject the jets at the center of the clumpy disk,\
  we define the atmosphere gas density by the peak density of\
  the isothermal disk on the mid-plane $z=0$ (i.e. $\rho^{\text{peak}}_{\text{isoDisk}}$),\
  as opposed to the \textit{clumpy} density around the jet source.

@@

 \begin{figure}%
      \captionsetup[subfigure]{labelformat=simple}
      \begin{subfigure}[b]{0.2\linewidth}%
         \includegraphics[height=10cm]{figures/fig__jetI5+ismSeed3-45deg-a}%
         \caption{}%
         \label{fig__jetI5+ismSeed3-45deg-a}%
      \end{subfigure}%
      \hspace{10pt}%
      \begin{subfigure}[b]{0.2\linewidth}%
         \includegraphics[height=10cm]{figures/fig__jetI5+ismSeed3-45deg-b}%
         \caption{}%
         \label{fig__jetI5+ismSeed3-45deg-b}%
      \end{subfigure}%
      \hspace{-2pt}%
      \begin{subfigure}[b]{0.2\linewidth}%
         \includegraphics[height=10cm]{figures/fig__jetI5+ismSeed3-45deg-c}%
         \caption{}%
         \label{fig__jetI5+ismSeed3-45deg-c}%
      \end{subfigure}%
      \hspace{-2pt}%
      \begin{subfigure}[b]{0.2\linewidth}%
         \includegraphics[height=10cm]{figures/fig__jetI5+ismSeed3-45deg-d}%
         \caption{}%
         \label{fig__jetI5+ismSeed3-45deg-d}%
     \end{subfigure}%
      \caption{
             The slices of pressure (top), temperature (middle), and number density (bottom)\
             at the end of simulation $t=12.39$ Myr.\
             The slices pass through a bipolar jet source injecting along $z=-y$ direction\
             for a duration $t=0$--$1.2$ Myr.\
             Comparison between the clumpy (Fig. \ref{fig__jetI5+ismSeed3-45deg-a})\
             and the smooth disks (Fig. \ref{fig__jetI5+ismSeed3-45deg-b}) in a stratified atmosphere\
             shows that the initial density distribution of the dense disk has an insignificant effect\
             on the overall dynamics of bubbles. However, the outermost bubbles arising from the smooth disk\
             in an uniform atmosphere (Fig. \ref{fig__jetI5+ismSeed3-45deg-c}) is nearly spherical,\
             suggesting that the stratification facilitates the elongation of the outermost bubbles significantly.\
             Also, Fig. \ref{fig__jetI5+ismSeed3-45deg-c} and \ref{fig__jetI5+ismSeed3-45deg-d}\
             reveal that the development of the innermost bubbles\
             is always associated with the disk. Also, without the disk (Fig. \ref{fig__jetI5+ismSeed3-45deg-d}),\
             the outermost bubbles and the turbulent plasma would be oblique,
             indicating that the dense disk is crucial for the production of symmetric Galactic bubbles.\ {\color{red} KY: Consider reducing the caption as it is essentially the same as the text.}
      }
      \label{fig__jetI5+ismSeed3-45deg}
 \end{figure}%

@@

\begin{figure*}
  \includegraphics[width=\linewidth]{figures/fig__GammaRay_100e9_1e6_angle_000.png}
  \caption{The observed (left; \citealt{Selig2015}) and simulated (right) photon flux\
           in the energy bin $76.8-153.6$ GeV.\
           Note that the left panel is the\
           photon flux of the diffuse component reconstructed by the D$^3$PO\
           algorithm \citep{Selig2015} that analyzes\
           the photon data from the \textit{Fermi} Large Area Telescope \citep{Atwood2009}\
           and removes the contribution from point-like component.
           The red arrow at the center of the right\
           panel depicts the direction of the bipolar jet, constantly ejecting at an angle of $45^{\circ}$\
           to the disk normal in 1.2 Myr.
  }
  \label{fig__gammaRay-map}
\end{figure*}
